# Instrumental Variable {#iv}

## IV without covariates X

**Goal**: Estimate the causal effect of $A$ on $Y$, especially when:

- A is "hopelessly" confounded
- A can never be randomized

**Idea**: Use "external" random variation in the data to "extract" randomness out of confounded A

**Controversy**: It's "impossible" to find such Z.

**Set up:** 

- $Z \in {0,1}$, Instrumental Variable
- $A \in {0,1}$, Treatment
- $Y \in \mathbb{R}$, Outcome

**Potential Outcome:**

- $A_i^{(Z=1)}$: potential treatment when $Z=1$
- $A_i^{(Z=0)}$: potential treatment when $Z=0$
- $Y_i^{(Z=1, A=1)}$,$Y_i^{(Z=0, A=1)}$,$Y_i^{(Z=1, A=0)}$,$Y_i^{(Z=0, A=0)}$
- $Y_i^{(z, a)}$ is the potential outcome if unit $i$ received $z$ and $a$

**Assumptions:**

- *A1*: SUTVA 
$$A_i = Z_iA_i(1) + (1-Z_i)A_i(0)$$
$$Y_i = Z_iY_i^{(1, A_i(1))} + (1-Z_i)Y_i^{(0, A_i(0))}$$
- *A2*: $Z \perp Y^{(z,a)}, A^{(z)}$, I.V. is randomized
- *A3*: $0 < \mathbb{P}(Z_i = 1) <1$
- *A4*: $\mathbb{E}[A(1) - A(0)] \neq 0$, I.V.(Z) has casual effect on A
- *A5*: $Y_i^{(Z=1, a)}=Y_i^{(Z=0, a)} = Y_i^{(a)}$
- *A6*: $A^{(1)} \geq A^{(0)}$, monotoncity

Notes on the assumptions:

- If Z was completely randomized like in a RCT, A1-A3 are plausible.
- A4 can be checked from data if Z was randomized
- A5 can never be fully verified with data!! We can observe $Y_i^{(1,a)}$,$Y_i^{(0,a)}$ at the same time.
- A6 means one-sided complicance exists in the experiment design. One-sided complicance: $A_i(0) = 0$, i.e $A_i(0) = 1$ will never happen.


Accoding to this paper, Angrist, Imbens and Rubin (1996) "Identification of Causal Effects Using Instrumental Variables", we can divide the subjects into four categories:

- *G1*: $A_i(1) =1,A_i(0) =1$: Always takers (doesn't exist for one-sided compliance)
- *G2*: $A_i(1) =0,A_i(0) =0$: Never takers
- *G3*: $A_i(1) =1,A_i(0) =0$: Compliers
- *G4*: $A_i(1) =0,A_i(0) =1$: Defiers

*G1-4* partitions the population in a non-overlapping manner. If monotonicity holds, then defiers cannot exists.

**Theorem:** If A1-A6 held, then 
\begin{align*}\mathbb{E}[Y_i(1) - Y_i(0) | \,i\text{ is complier }] &= \mathbb{E}[Y_i(1) - Y_i(0) | A_i(1)=1, A_i(0)=1] \\
&= \frac{\mathbb{E}(Y_i | Z_i=1)-\mathbb{E}(Y_i | Z_i=0)}{\mathbb{E}(A_i | Z_i=1)- \mathbb{E}(A_i | Z_i=0)}
\end{align*}

Note that the estimand identifies the ATE, but among a group of people (compliers) and we don't know who they are!

**proof:**

First, let us look at the numerator.

\begin{align*}
&\,\,\,\,\,\,\,\,E(Y_i |Z_i=1) - E(Y_i|Z_i=0) \\
&=^{A_1} E(Z_iY_i^{(1,A_i(1))}+(1-Z_i)Y_i^{(0,A_i(0))} |Z_i=1  )-\\
&\,\,\,\,\,\,\,\,E(Z_iY_i^{(1,A_i(1))}+(1-Z_i)Y_i^{(0,A_i(0))} |Z_i=0  ) \\
&= E(Y_i^{(1, A_i(1))}|Z_i=1) - E(Y_i^{(0, A_i(0))}|Z_i=0) \\
&=^{A_2} E(Y_i^{(1, A_i(1))}) - E(Y_i^{(0, A_i(0))}) \\
&= E[ (Y_i^{(1, A_i(1))}) - E(Y_i^{(0, A_i(0))})I(A_i(1)=1,A_i(0)=0)+\\
&\,\,\,\,\,\,\,(Y_i^{(1, A_i(1))}) - E(Y_i^{(0, A_i(0))})I(A_i(1)=1,A_i(0)=1)+\\
&\,\,\,\,\,\,\,(Y_i^{(1, A_i(1))}) - E(Y_i^{(0, A_i(0))})I(A_i(1)=0,A_i(0)=0)]\\
& = E[ (Y_i^{(1, A_i(1))}) - E(Y_i^{(0, A_i(0))})I(A_i(1)=1,A_i(0)=0) ] \\
& = E[ (Y_i^{(1)}) - E(Y_i^{(0)})I(A_i(1)=1,A_i(0)=0) ]
\end{align*}

Then, we look at the denominator.
\begin{align*}
&\,\,\,\,\,\,\,\,E[A_i |Z_i=1] - E[A_i|Z_i=0]\\
&= E[Z_iA_i(1)+(1-Z_i)A_i(0)|Z_i=1] -\\
&\,\,\,\,\,\,\,\,E[Z_iA_i(1)+(1-Z_i)A_i(0)|Z_i=0]\\
&= E[A_i(1) |Z_i=1] - E[A_i(0)|Z_i=0]\\
&= E[A_i(1) ] - E[A_i(0)] \\
&= E[A_i(1)  - A_i(0)]
\end{align*}

Thus,
\begin{align*}
&\,\,\,\,\,\,\,\,E[Y_i(1) - Y_i(0) | A_i(1)=1, A_i(0)=1] \\
&= \frac{E[(Y_i(1) - Y_i(0))I(A_i(1)=1, A_i(0)=1)]}{E[I(A_i(1)=1, A_i(0)=1)]} \\
&= \frac{E[(Y_i(1) - Y_i(0))I(A_i(1)=1, A_i(0)=1)]}{E[A_i(1) - A_i(0)]}
\end{align*}

**How to estmate?**

**First method:**We can use the sample mean to estimate the terms in the theorem, i.e.
$$\hat E(Y_i |Z_i=1) = \frac{\sum_{i=1}^{n}Y_iZ_i}{\sum_{i=1}^{n}Z_i} $$

**Theorem: ** $\hat E(Y_i |Z_i=1) \overset{p}{\to} E(Y_i |Z_i=1)$

**proof:** 

$$\frac{1}{n}\sum_{i=1}^{n}Y_iZ_i \overset{p}{\to} E[Y_iZ_i]$$
$$\frac{1}{n}\sum_{i=1}^{n}Z_i \overset{p}{\to} E[Z_i]$$
Thus, by using Slutsky's theorem:

$$\frac{\sum_{i=1}^{n}Y_iZ_i}{\sum_{i=1}^{n}Z_i} \overset{p}{\to} \frac{E[Y_iZ_i]}{E[Z_i]} = E[Y_i|Z_i=1]$$

**Theorem:** 

$$\hat{\tau}_w = \frac{\hat{E}_{YZ}-\hat{E}_{Y(1-Z)}}{\hat{E}_{AZ}-\hat{E}_{A(1-Z)}} \overset{p}{\to} \tau = \frac{{E}_{YZ}-{E}_{Y(1-Z)}}{{E}_{AZ}-{E}_{A(1-Z)}}$$
and

$$\sqrt{\hat{\tau}_w - \tau} \to N(0, E\bigg[ \frac{\frac{Y_i(1)^2}{p}+\frac{Y_i(0)^2}{1-p}}{\tau_A^2} \bigg] - \tau^2),$$
where $p = \mathbb{P}(Z_i=1)$ and $\tau_A = E[A_i |Z_i=1]-E[A_i |Z_i=0]$

**Second method:** Two Stage Least Square

```{r, eval=FALSE}
lm.fit1 <- lm(A~Z)
Ahat <- predict(lm.fit1)
lm.fit2 <- lm(Y~Ahat)
tau_tsls <- coef(lm.fit2)[2]
```

**Theorem:** If $Z$ is binary, then $\hat{\tau}_{TSLS} = \hat{\tau}_w$

## IV with covariates X

**Assumptions:**

- *A1*: SUTVA 
$$A_i = Z_iA_i(1) + (1-Z_i)A_i(0)$$
$$Y_i = Z_iY_i^{(1, A_i(1))} + (1-Z_i)Y_i^{(0, A_i(0))}$$
- *A2*: $Z_i \perp A_i(1),A_i(0), Y_i^{(1, A_i(1))},Y_i^{(0, A_i(0))} | X_i$
- *A3*: $0 < \mathbb{P}(Z_i = 1 |X_i=x) <1\,\,\forall x$
- *A4*: $\mathbb{E}[A(1) - A(0)|X_i=x] \neq 0\,\,\forall x$
- *A5*: $Y_i^{(Z=1, a)}=Y_i^{(Z=0, a)} = Y_i^{(a)}$
- *A6*: $A^{(1)} \geq A^{(0)}$, monotoncity

**Theorem:** Under assumptions *A1-A6*, The ATE among compliers:

$${E}[Y_i(1) - Y_i(0) | A_i(1)=1, A_i(0)=1]=\frac{E_X\{E[Y_i|Z_i=1,X_i]-E[Y_i|Z_i=0,X_i]\}}{E_X\{E[A_i|Z_i=1,X_i]-E[A_i|Z_i=0,X_i]\}}$$

**Strategy of estimating with Xs:**

**1:Two stage least square**

```{r, eval=FALSE}
lm.fit1 <- lm(A~Z+X)
Ahat <- predict(lm.fit1)
lm.fit2 <- lm(Y~Ahat+X)
tau_tsls <- coef(lm.fit2)
```

**Property of $\hat \tau_{TSLS}$**

- $\hat \tau_{TSLS}\to\tau$
- $\sqrt{n}(\hat \tau_{TSLS}-\tau) \to N(0, V_{TSLS})$

**2:Matching (non-parametric)**

```{r, eval=FALSE}
lm.fit1 <- lm(Y~Z + matched set indicators)
lm.fit2 <- lm(Y~Z + matched set indicators)
```

Take the ratio of the coef estimate between $Y\sim Z+.$ and $A\sim Z+.$

**Property of $\hat \tau_{match}$**

- $\hat \tau_{match}\to\tau$
- $\sqrt{n}(\hat \tau_{match}-\tau) \to N(0, V_{match})$
- $V_{match} \geq V_{TSLS}$

**3:IPW OR DR**

## New topics in I.V.

- Multiple IVs 
- Privacy and IV


